Galois Characterization of Function Fields of \\ Transcendence Degree $1$ over Number Fields

### 2017-08-15 Master Thesis/synopsis.tex
Roots of unity are contained in absolute inertia field

Let $T$ be the absolute inertia group of $K$ and $V\lhd T$ its absolute ramification group. By Theorem \ref{theorem-sylowgroups-inertiagroup}, $V$ is the $p$-Sylow group of $T$ and the $\ell$-Sylow group of $T/V$ is isomorphic to $\mathbb{Z}_\ell$. Therefore, by \cite{Se64} Ch. I, §3 Proposition 15:
\[ \operatorname{cd}_\ell(T)\leq \operatorname{cd}_\ell(V) + \operatorname{cd}_\ell(\mathbb{Z}_\ell) = 1 \]
By Theorem \ref{theorem-absoluteinertia}, we have an isomorphism $G_K/T\cong G_k$. Furthermore, we have $\operatorname{cd}_\ell(G_K/T)=\operatorname{cd}_\ell(k)\leq 2$ by requirement,
thus the following Hochschild-Serre spectral sequence degenerates on $E_2$:
\[E_2^{i,j}=\mathrm{H}^i(G_K/T, \mathrm{H}^j(T, \mathbb{Z}/\ell\mathbb{Z}(2)))\Longrightarrow\mathrm{H}^{i+j}(G_K, \mathbb{Z}/\ell\mathbb{Z}(2))\]
It yields an isomorphism $\mathrm{H}^2(k, \mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(2)))\cong \mathrm{H}^3(K, \mathbb{Z}/\ell\mathbb{Z}(2))$. Since $\mu_{\ell}$ is contained in the absolute inertia field of $K$, $T$ acts trivially on $\mathbb{Z}/\ell\mathbb{Z}(1)= \mu_{\ell}$ and $\mathbb{Z}/\ell\mathbb{Z}(2)=\mu_{\ell}^{\otimes 2}$, hence:
\begin{align*}
\mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(2)) &\cong \mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(1)) \\
&\cong \operatorname{Hom}_\text{cont.}(T, \mathbb{Z}/\ell\mathbb{Z}(1)) \\
&\cong \operatorname{Hom}_\text{cont.}(T/V, \mathbb{Z}/\ell\mathbb{Z}(1))
\end{align*}
By Theorem \ref{theorem-sylowgroups-inertiagroup}, $T/V$ is the free pro-$\mathcal{C}$ group of rank $1$ where $\mathcal{C}$ is the class of all finite groups with order prime to $p$. Therefore, $\operatorname{Hom}_\text{cont.}(T/V, \mathbb{Z}/\ell\mathbb{Z}(1))\cong \mathbb{Z}/\ell\mathbb{Z}(1)$. The second isomorphism follows from Proposition \ref{prop:kummer-theory}
\end{proof}

@@

\begin{proof}
\begin{enumerate}[(i)]
\item Consider the absolute inertia field $L = K_P^\text{t}$ of $K_P$ and let $w$ be the unique extension of $v_P$ in $L$. By Theorem \ref{theorem-absoluteinertia} (i), $w$ is discrete as well and we have $k(w) = k(v_P)^\text{s}$. Observe the following exact sequence:
\[ 1 \longrightarrow U_w/n \longrightarrow L^\times /n\longrightarrow \mathbb{Z}/n\mathbb{Z}\longrightarrow 0 \]
Since $k(w)$ is separably closed, with $\operatorname{char} k(w)=0$ and Hensel's lemma, we can see that $U_w$ is a divisible group, hence we get an isomorphism $L^\times /n\cong \mathbb{Z}/n\mathbb{Z}$ induced by $w$. Therefore, the canonical map $K^\times/n\to L^\times/n$ can be identified with the map $K^\times/n \to\mathbb{Z}/n\mathbb{Z}$ induced by $v_P$. Now the following commutative diagram with exact columns gives us the result:
\[ \begin{tikzcd}
1 \ar[d] & 1 \ar[d]\\
U_{P'}/n \ar[r, dashed]\ar[d] & U_P/n\ar[d]\\
K'^\times /n \ar[r, "\varphi_n"]\ar[d] & K^\times/n\ar[d]\\
L'^\times /n \ar[r, "\varphi_n"'] &  L^\times /n
\end{tikzcd}\]
Note that $(K^\text{t}_P)' = K'^{\text{t}}_{P'}$ as proven in Lemma \ref{3.2}.
\item Consider the field extension $K_P/K$. By Remark \ref{remark-localphi-extension}, we can assume $K=K_P$ without loss of generality. Since $\operatorname{char}k(P)=0$, by Hensel's lemma, the map $r_P/n: U_P/n\to k(P)^\times/n$ is an isomorphism and the composition
\[\begin{tikzcd}
k(P)^\times / n\ar[rr, "(r_P/n)^{-1}"] && U_P/n \ar[rr, hook] && K_P^\times/n
\end{tikzcd}\]
coincides with the inflation $\operatorname{inf}: \mathrm{H}^1(k(P), \mu_n) \to \mathrm{H}^1(K_P,\mu_n)$. The same holds for $P'$ due to Lemma \ref{3.2}. Therefore, the assertion follows from the commutativity of the following diagram:
\[ \begin{tikzcd}
k(P')^\times/n \ar[r, "\cong"] \ar[d, "\varphi_P^\times/n"'] & \mathrm{H}^1(k(P'),\mu_n) \ar[r, "\mathrm{inf}"] \ar[d]& \mathrm{H}^1(K'_{P'}, \mu_n) \ar[d, "\varphi_n"] \\
k(P)^\times/n \ar[r, "\cong"'] & \mathrm{H}^1(k(P),\mu_n)  \ar[r, "\mathrm{inf}"'] & \mathrm{H}^1(K_{P}, \mu_n)
\end{tikzcd}  \qedhere\]
\end{enumerate}
\end{proof}

>>>>

Let $T$ be the absolute inertia group of $K$ and $V\lhd T$ its absolute ramification group. By Theorem \ref{theorem-sylowgroups-inertiagroup}, $V$ is the $p$-Sylow group of $T$ and the $\ell$-Sylow group of $T/V$ is isomorphic to $\mathbb{Z}_\ell$. Therefore, by \cite{Se64} Ch. I, §3 Proposition 15:
\[ \operatorname{cd}_\ell(T)\leq \operatorname{cd}_\ell(V) + \operatorname{cd}_\ell(\mathbb{Z}_\ell) = 1 \]
By Theorem \ref{theorem-absoluteinertia}, we have an isomorphism $G_K/T\cong G_k$. Furthermore, we have $\operatorname{cd}_\ell(G_K/T)=\operatorname{cd}_\ell(k)\leq 2$ by requirement,
thus the following Hochschild-Serre spectral sequence degenerates on $E_2$:
\[E_2^{i,j}=\mathrm{H}^i(G_K/T, \mathrm{H}^j(T, \mathbb{Z}/\ell\mathbb{Z}(2)))\Longrightarrow\mathrm{H}^{i+j}(G_K, \mathbb{Z}/\ell\mathbb{Z}(2))\]
It yields an isomorphism $\mathrm{H}^2(k, \mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(2)))\cong \mathrm{H}^3(K, \mathbb{Z}/\ell\mathbb{Z}(2))$. Let $K^\text{t}/K$ be the absolute inertia field of $K$ and $w\in\operatorname{Spv}(K^\text{t})$ the extension of $v$. Since $k(w)$ is separably closed by Theorem \ref{theorem-absoluteinertia}, $\mu_{\ell}$ is contained in $K^\text{t}$ by Hensel's lemma. Therefore, $T = G_{K^\text{t}}$ acts trivially on $\mathbb{Z}/\ell\mathbb{Z}(1)= \mu_{\ell}$ and $\mathbb{Z}/\ell\mathbb{Z}(2)=\mu_{\ell}^{\otimes 2}$, hence:
\begin{align*}
\mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(2)) &\cong \mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(1)) \\
&\cong \operatorname{Hom}_\text{cont.}(T, \mathbb{Z}/\ell\mathbb{Z}(1)) \\
&\cong \operatorname{Hom}_\text{cont.}(T/V, \mathbb{Z}/\ell\mathbb{Z}(1))
\end{align*}
By Theorem \ref{theorem-sylowgroups-inertiagroup}, $T/V$ is the free pro-$\mathcal{C}$ group of rank $1$ where $\mathcal{C}$ is the class of all finite groups with order prime to $p$. Therefore, $\operatorname{Hom}_\text{cont.}(T/V, \mathbb{Z}/\ell\mathbb{Z}(1))\cong \mathbb{Z}/\ell\mathbb{Z}(1)$. The second isomorphism follows from Proposition \ref{prop:kummer-theory}
\end{proof}

@@

\begin{proof}
\begin{enumerate}[(i)]
\item Consider the absolute inertia field $L = K_P^\text{t}$ of $K_P$ and let $w$ be the unique extension of $v_P$ in $L$. By Theorem \ref{theorem-absoluteinertia} (i), $w$ is discrete as well and we have $k(w) = k(v_P)^\text{s}$. Observe the following exact sequence:
\[ 1 \longrightarrow U_w/n \longrightarrow L^\times /n\longrightarrow \mathbb{Z}/n\mathbb{Z}\longrightarrow 0 \]
Since $k(w)$ is separably closed, with Hensel's lemma, we can see that $U_w$ is a divisible group, hence we get an isomorphism $L^\times /n\cong \mathbb{Z}/n\mathbb{Z}$ induced by $w$. Therefore, the canonical map $K^\times/n\to L^\times/n$ can be identified with the map $K^\times/n \to\mathbb{Z}/n\mathbb{Z}$ induced by $v_P$. Now the following commutative diagram with exact columns gives us the result:
\[ \begin{tikzcd}
1 \ar[d] & 1 \ar[d]\\
U_{P'}/n \ar[r, dashed]\ar[d] & U_P/n\ar[d]\\
K'^\times /n \ar[r, "\varphi_n"]\ar[d] & K^\times/n\ar[d]\\
L'^\times /n \ar[r, "\varphi_n"'] &  L^\times /n
\end{tikzcd}\]
Note that $(K^\text{t}_P)' = K'^{\text{t}}_{P'}$ as proven in Lemma \ref{3.2}.
\item Consider the field extension $K_P/K$. By Remark \ref{remark-localphi-extension}, we can assume $K=K_P$ without loss of generality. By Hensel's lemma, the map $r_P/n: U_P/n\to k(P)^\times/n$ is an isomorphism and the composition
\[\begin{tikzcd}
k(P)^\times / n\ar[rr, "(r_P/n)^{-1}"] && U_P/n \ar[rr, hook] && K_P^\times/n
\end{tikzcd}\]
coincides with the inflation $\operatorname{inf}: \mathrm{H}^1(k(P), \mu_n) \to \mathrm{H}^1(K_P,\mu_n)$. The same holds for $P'$ due to Lemma \ref{3.2}. Therefore, the assertion follows from the commutativity of the following diagram:
\[ \begin{tikzcd}
k(P')^\times/n \ar[r, "\cong"] \ar[d, "\varphi_P^\times/n"'] & \mathrm{H}^1(k(P'),\mu_n) \ar[r, "\mathrm{inf}"] \ar[d]& \mathrm{H}^1(K'_{P'}, \mu_n) \ar[d, "\varphi_n"] \\
k(P)^\times/n \ar[r, "\cong"'] & \mathrm{H}^1(k(P),\mu_n)  \ar[r, "\mathrm{inf}"'] & \mathrm{H}^1(K_{P}, \mu_n)
\end{tikzcd}  \qedhere\]
\end{enumerate}
\end{proof}
